---
title: "EEB313 term project"
Author: "Yuan Xiao, Advait Sridhar, Xi Nan"
date: 14/12/2021
output: html_notebook
---

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lme4)
library(lmerTest)
library(ggalt)
```

```{r setup}
knitr::opts_knit$set(root.dir="~/Desktop/School/EEB313/Term project")
```

```{r}
ontario <- read.csv("Ontario.csv")
erie <- read.csv("Erie.csv")
ontario
erie
```

## Transform data -- gather species names into one column and create a separate abundance column for each species ##
```{r}
ontario_long<- ontario %>%
  gather(key= species, value= species_catch, Lake.Whitefish, Yellow.Perch, Freshwater.Drum, Walleye) 

erie_long<- erie %>%
  gather(key= species, value= species_catch, Lake.Whitefish, Yellow.Perch, Freshwater.Drum, Walleye) 
```

Get actual abundance of each species by dividing species_catch, which is given in thousands of pounds, by the average weight of each species and multiplying by 1000.
```{r}
#Create new columns for the abundance of each species
ontario_long <- ontario_long %>%
  spread(key= species, value= species_catch) %>%
  mutate(Freshwater.Drum_catch= (Freshwater.Drum*1000/5)) %>%
  mutate(Lake.Whitefish_catch= (Lake.Whitefish*1000/3)) %>%
  mutate(Walleye_catch= (Walleye*1000/2)) %>%
  mutate(Yellow.Perch_catch= (Yellow.Perch*1000/0.75)) %>%
  gather(key= species, value= species_catch, Freshwater.Drum_catch, Lake.Whitefish_catch, Walleye_catch, Yellow.Perch_catch)
ontario_long
```

```{r}
#Create new columns for abundance of each species
erie_long <- erie_long %>%
  spread(key= species, value= species_catch) %>%
  mutate(Freshwater.Drum_catch= (Freshwater.Drum*1000/5)) %>%
  mutate(Lake.Whitefish_catch= (Lake.Whitefish*1000/3)) %>%
  mutate(Walleye_catch= (Walleye*1000/2)) %>%
  mutate(Yellow.Perch_catch= (Yellow.Perch*1000/0.75)) %>%
  gather(key= species, value= species_catch, Freshwater.Drum_catch, Lake.Whitefish_catch, Walleye_catch, Yellow.Perch_catch)
erie_long
```

```{r}
# combine both datasets into one 
lakes_long<- rbind(ontario_long, erie_long) 
lakes_long$LakeName <- as.factor(lakes_long$LakeName)
lakes_long$species <- as.factor(lakes_long$species)
lakes_long
```

## Graph chlorophyll and temperature over time ##
```{r}
#plot chlorophyll level line graph of the two lakes
lakes_long %>% 
  ggplot(aes(x=Year, y=`Avg..Chl.`, color=LakeName))+
  geom_line()+
  ggtitle("Yearly Chlorophll Levels of Lake Ontario and Lake Erie from 1995 to 2011")

#plot temperature line graph of two lakes 
lakes_long %>% 
  ggplot(aes(x=Year, y=`Avg..Temp.`, color=LakeName))+
  geom_line()+
  ggtitle("Yearly Temperature Levels of Lake Ontario and Lake Erie from 1995 to 2011")
```

## Graph abundance over time ##

```{r}
#plot species abundance line graph of Lake Ontario
ontario_long %>% 
  ggplot(aes(x=Year, y= log10(species_catch), color=species))+
  geom_line()+
  ggtitle("Abundance of Four Species on Lake Ontario from 1995 to 2011")

#plot species abundance line graph of Lake Erie
erie_long %>% 
  ggplot(aes(x=Year, y= log10(species_catch), color=species))+
  geom_line()+
  ggtitle("Abundance of Four Species on Lake Erie from 1995 to 2011")  
```


## Separate linear models for temperature and chlorophyll ##

```{r}
# Model for chlorophyll and abundance
lakes_lmChl <- lmer(log10(species_catch) ~ Avg..Chl.*species + (1|LakeName), data= lakes_long, REML=FALSE)
summary(lakes_lmChl)
```

```{r}
# Test assumptions of linear model
lm_lakes_long_chl <- lm(log10(species_catch) ~ Avg..Chl. * species, data = lakes_long) #creating linear model with lm() because plot function doesn't work on lmer objects
plot(lm_lakes_long_chl, 1:2) #testing assumptions for Avg..Chl.*species interaction
```

```{r}
#ANOVA table
lakes_anova<- Anova(lmer(log10(species_catch) ~ Avg..Chl.*species + (1|LakeName), data= lakes_long), type=3)
lakes_anova
```

```{r}
#aov object for Tukey test
lakes_aovChl<- aov(log10(species_catch) ~ Avg..Chl.*species, data= lakes_long)
summary(lakes_aovChl)
```

```{r}
#Tukey test
Tukey<- TukeyHSD(lakes_aovChl)
Tukey
```

```{r}
# graph for tukey test
par(mar=c(5, 19, 5, 5))
plot(Tukey, las=1)
```

```{r}
# Add model predictions to data frame
lakes_long <- lakes_long %>%
  mutate(pred_Chl= predict(lakes_lmChl))

# plot for chlorophyll
ggplot(lakes_long, aes(x = Avg..Chl., y = log10(species_catch), colour = species)) +
  geom_abline(aes(intercept = 3.72, slope = 80.65), size = 2) +       # Add the intercept and estimate for chlorophyll from the model
  geom_line(aes(y= pred_Chl), size = 1) +
  geom_point(size = 2) + 
  facet_wrap(~species) +
  ylab("log10(Species abundance)") +
  xlab("Average chlorophyll concentration (mg/L)")
```

```{r}
# Model for temperature and abundance
lakes_lmTemp <- lmer(log10(species_catch) ~ Avg..Temp.*species + (1|LakeName), data= lakes_long, REML=FALSE)
summary(lakes_lmTemp)
```

```{r}
# Test assumptions for linear model
lm_lakes_long_Temp <- lm(log10(species_catch) ~ Avg..Chl. * species, data = lakes_long) #creating linear model with lm() because plot function doesn't work on lmer objects
plot(lm_lakes_long_Temp, 1:2) 
```

```{r}
# Add model predictions to the data frame
lakes_long <- lakes_long %>%
  mutate(pred_Temp= predict(lakes_lmTemp))

# Graph the results of the temperature model
ggplot(lakes_long, aes(x = Avg..Temp., y = log10(species_catch), colour = species)) +
  geom_abline(aes(intercept = 7.339916, slope = -0.31024), size = 2) +       
  #geom_smooth(method="lm", size = 1) +
  geom_line(aes(y= pred_Temp), size = 1) +
  geom_point(size = 2) + 
  facet_wrap(~species) +
  ylab("log10(Species abundance)") +
  xlab("Average surface temperature (°C)")
```

```{r}
# Anova table for temperature model 
lakes_anovatemp <- Anova(lmer(log10(species_catch) ~ Avg..Temp.*species + (1|LakeName), data= lakes_long), type=3)
lakes_anovatemp
```

```{r}
# create aov object for the Tukey test
lakes_aovTemp<- aov(log10(species_catch) ~ Avg..Temp.*species, data= lakes_long)
summary(lakes_aovTemp)
```

```{r}
# Tukey test
Tukey2 <- TukeyHSD(lakes_aovTemp)
Tukey2
```

```{r}
# graph for tukey test
par(mar=c(5, 19, 5, 5))
plot(Tukey2, las=1)
```



